---
title: "SLR: Matrix representation"
subtitle: "cont'd"
author: "Prof. Maria Tackett"
date: "2024-09-10"
date-format: "MMM DD, YYYY"
footer: "[🔗 STA 221 - Fall 2024](https://sta221-fa24.netlify.app)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r setup}
#| include: false

library(countdown)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%",
  fig.align = "center"
)
```

## Announcements

-   Lab 01 due on **Thursday, September 12 at 11:59pm**

    -   Push work to GitHub repo

    -   Submit final PDF on Gradescope + mark pages for each question

-   HW 01 will be assigned on Thursday

## Topics

-   Matrix representation of simple linear regression
    -   Model form
    -   Least square estimate
    -   Predicted (fitted) values
    -   Residuals

```{r packages}
#| echo: false
#| message: false
#| warning: false

# load packages
library(tidyverse)   # for data wrangling and visualization
library(tidymodels)  # for modeling (includes broom, yardstick, and other packages)
library(openintro)   # for the duke_forest dataset
library(scales)      # for pretty axis labels
library(knitr)       # for pretty tables
library(patchwork)   # arrange plots

# set default theme for ggplot2
ggplot2::theme_set(ggplot2::theme_bw())
```

# Matrix representation of simple linear regression

## SLR in matrix form {.midi}

Suppose we have $n$ observations, a quantitative response variable, and a single predictor$$
\underbrace{
\begin{bmatrix}
y_1 \\
\vdots \\
y_n
\end{bmatrix} }_
{\mathbf{y}} \hspace{3mm}
= 
\hspace{3mm}
\underbrace{
\begin{bmatrix}
1 &x_1 \\
\vdots &  \vdots \\
1 &  x_n
\end{bmatrix}
}_{\mathbf{X}}
\hspace{2mm}
\underbrace{
\begin{bmatrix}
\beta_0 \\
\beta_1
\end{bmatrix}
}_{\boldsymbol{\beta}}
\hspace{3mm}
+
\hspace{3mm}
\underbrace{
\begin{bmatrix}
\epsilon_1 \\
\vdots\\
\epsilon_n
\end{bmatrix}
}_\boldsymbol{\epsilon}
$$

<br>

-   $\mathbf{y}$: $n\times 1$ vector of responses
-   $\mathbf{X}$: $n \times 2$ design matrix
-   $\boldsymbol{\beta}$: $2 \times 1$ vector of coefficients
-   $\boldsymbol{\epsilon}$: $n \times 1$ vector of error terms

## Minimize sum of squared residuals

**Goal**: Find $\boldsymbol{\beta} = \begin{bmatrix}\beta_0 \\ \beta_1 \end{bmatrix}$ that minimizes the sum of squared residuals $$
\begin{aligned}
SSR = \sum_{i=1}^n e_i^2 = \mathbf{e}^T\mathbf{e} &= (\mathbf{y} - \mathbf{X}\boldsymbol{\beta})^T(\mathbf{y} - \mathbf{X}\boldsymbol{\beta}) \\[10pt]
\end{aligned}
$$

## Minimize sum of squared residuals

$$
\begin{aligned}
\mathbf{e}^T\mathbf{e} &= (\mathbf{y} - \mathbf{X}\boldsymbol{\beta})^T(\mathbf{y} - \mathbf{X}\boldsymbol{\beta}) \\[10pt]
&\class{fragment}{= (\mathbf{y}^T - \boldsymbol{\beta}^T\mathbf{X}^T)(\mathbf{y} - \mathbf{X}\boldsymbol{\beta})}\\[10pt]
&\class{fragment}{=\mathbf{y}^T\mathbf{y} - \mathbf{y}^T\mathbf{X}\boldsymbol{\beta} - \boldsymbol{\beta}^T\mathbf{X}^T\mathbf{y} + \boldsymbol{\beta}^T\mathbf{X}^T\mathbf{X}\boldsymbol{\beta}}\\[10pt]
\end{aligned}
$$

## Minimize sum of squared residuals

$$
\begin{aligned}
\mathbf{e}^T\mathbf{e} &= (\mathbf{y} - \mathbf{X}\boldsymbol{\beta})^T(\mathbf{y} - \mathbf{X}\boldsymbol{\beta}) \\[10pt]
&= (\mathbf{y}^T - \boldsymbol{\beta}^T\mathbf{X}^T)(\mathbf{y} - \mathbf{X}\boldsymbol{\beta})\\[10pt]
&=\mathbf{y}^T\mathbf{y} - \mathbf{y}^T\mathbf{X}\boldsymbol{\beta} - \boldsymbol{\beta}^T\mathbf{X}^T\mathbf{y} + \boldsymbol{\beta}^T\mathbf{X}^T\mathbf{X}\boldsymbol{\beta}\\[10pt]
&=\mathbf{y}^T\mathbf{y} - 2\boldsymbol{\beta}^T\mathbf{X}^T\mathbf{y} + \boldsymbol{\beta}^T\mathbf{X}^T\mathbf{X}\boldsymbol{\beta}
\end{aligned}
$$

## Minimize sum of squared residuals

The estimate of $\boldsymbol{\beta} = \begin{bmatrix}\beta_0 \\ \beta_1 \end{bmatrix}$ that minimizes SSR is the one such that

$$
\nabla_{\boldsymbol{\beta}} SSR = \nabla_{\boldsymbol{\beta}}( \mathbf{y}^T\mathbf{y} - 2\boldsymbol{\beta}^T\mathbf{X}^T\mathbf{y} + \boldsymbol{\beta}^T\mathbf{X}^T\mathbf{X}\boldsymbol{\beta}) = 0
$$

## Side note: Vector operations {.midi}

Let $\mathbf{x} = \begin{bmatrix}x_1 \\ x_2 \\ \vdots \\x_k\end{bmatrix}$be a $k \times 1$ vector and $f(\mathbf{x})$ be a function of $\mathbf{x}$.

. . .

Then $\nabla_\mathbf{x}f$, the **gradient** of $f$ with respect to $\mathbf{x}$ is

$$
\nabla_\mathbf{x}f = \begin{bmatrix}\frac{\partial f}{\partial x_1} \\ \frac{\partial f}{\partial x_2} \\ \vdots \\ \frac{\partial f}{\partial x_k}\end{bmatrix}
$$

## Side note: Vector operations

Th **Hessian** matrix, $\nabla_\mathbf{x}^2f$ is a $k \times k$ matrix of partial second derivatives

$$
\nabla_{\mathbf{x}}^2f = \begin{bmatrix} \frac{\partial^2f}{\partial x_1^2} & \frac{\partial^2f}{\partial x_1 \partial x_2} & \dots & \frac{\partial^2f}{\partial x_1\partial x_k} \\ 
\frac{\partial^2f}{\partial\ x_2 \partial x_1} & \frac{\partial^2f}{\partial x_2^2} & \dots & \frac{\partial^2f}{\partial x_2 \partial x_k} \\
\vdots & \vdots & \ddots & \vdots \\ 
\frac{\partial^2f}{\partial x_k\partial x_1} & \frac{\partial^2f}{\partial x_k\partial x_2} & \dots & \frac{\partial^2f}{\partial x_k^2} \end{bmatrix}
$$

## Side note: Vector operations

::: {.callout-note icon="false"}
## Proposition 1

Let $\mathbf{x}$ be a $k \times 1$ vector and $\mathbf{z}$ be a $k \times 1$ vector, such that $\mathbf{z}$ is not a function of $\mathbf{x}$ .

The gradient of $\mathbf{x}^T\mathbf{z}$ with respect to $\mathbf{x}$ is

$$
\nabla_\mathbf{x} \hspace{1mm} \mathbf{x}^T\mathbf{z} = \mathbf{z}
$$
:::

## Side note: Proposition 1

$$
\begin{aligned}
\mathbf{x}^T\mathbf{z} &= \class{fragment}{\begin{bmatrix}x_1 & x_2 & \dots &x_k\end{bmatrix}
 \begin{bmatrix}z_1 \\ z_2 \\ \vdots \\z_k\end{bmatrix}} \\[10pt]
 &\class{fragment}{= x_1z_1 + x_2z_2 + \dots + x_kz_k} \\
&\class{fragment}{= \sum_{i=1}^k x_iz_i}
\end{aligned}
$$

## Side note: Proposition 1 {.midi}

$$
\nabla_\mathbf{x}\hspace{1mm}\mathbf{x}^T\mathbf{z} = \class{fragment}{\begin{bmatrix}\frac{\partial \mathbf{x}^T\mathbf{z}}{\partial x_1} \\ \frac{\partial \mathbf{x}^T\mathbf{z}}{\partial x_2} \\ \vdots \\ \frac{\partial \mathbf{x}^T\mathbf{z}}{\partial x_k}\end{bmatrix}}  
= \class{fragment}{\begin{bmatrix}\frac{\partial}{\partial x_1} (x_1z_1 + x_2z_2 + \dots + x_kz_k) \\ \frac{\partial}{\partial x_2} (x_1z_1 + x_2z_2 + \dots + x_kz_k)\\ \vdots \\ \frac{\partial}{\partial x_k} (x_1z_1 + x_2z_2 + \dots + x_kz_k)\end{bmatrix}}
 = \class{fragment}{\begin{bmatrix} z_1 \\ z_2 \\ \vdots \\ z_k\end{bmatrix} = \mathbf{z}}
$$

## Side note: Vector + matrix operations

::: {.callout-note icon="false"}
## Proposition 2

Let $\mathbf{x}$ be a $k \times 1$ vector and $\mathbf{A}$ be a $k \times k$ matrix, such that $\mathbf{A}$ is not a function of $\mathbf{x}$ .

Then the gradient of $\mathbf{x}^T\mathbf{A}\mathbf{x}$ with respect to $\mathbf{x}$ is

$$
\nabla_\mathbf{x} \hspace{1mm} \mathbf{x}^T\mathbf{A}\mathbf{x} = (\mathbf{A}\mathbf{x} + \mathbf{A}^T \mathbf{x}) = (\mathbf{A} + \mathbf{A}^T)\mathbf{x}
$$

If $\mathbf{A}$ is symmetric, then

$$
(\mathbf{A} + \mathbf{A}^T)\mathbf{x} = 2\mathbf{A}\mathbf{x}
$$
:::

<center>Proof in HW 01</center>

## Find the least squares estimators

$$
\begin{aligned}
\nabla_{\boldsymbol{\beta}} SSR &= \nabla_{\boldsymbol{\beta}}( \mathbf{y}^T\mathbf{y} - 2\boldsymbol{\beta}^T\mathbf{X}^T\mathbf{y} + \boldsymbol{\beta}^T\mathbf{X}^T\mathbf{X}\boldsymbol{\beta})  \\[10pt]
& \class{fragment}{= \nabla_\boldsymbol{\beta} \hspace{1mm} \mathbf{y}^T\mathbf{y} - 2\nabla_\boldsymbol{\beta} \hspace{1mm} \boldsymbol{\beta}^T\mathbf{X}^T\mathbf{y} + \nabla_\boldsymbol{\beta} \hspace{1mm} \boldsymbol{\beta}^T\mathbf{X}^T\mathbf{X}\boldsymbol{\beta}} \\[10pt]
&\class{fragment}{= 0 - 2\mathbf{X}^T\mathbf{y} + 2\mathbf{X}^T\mathbf{X}\boldsymbol{\beta}}\class{fragment}{=0} \\[10pt]
&\class{fragment}{\Rightarrow \mathbf{X}^T\mathbf{X}\boldsymbol{\beta} = \mathbf{X}^T\mathbf{y}} \\[10pt]
&\class{fragment}{\Rightarrow   (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{X}\boldsymbol{\beta} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}} \\[10pt]
&\class{fragment}{\color{#993399}{\Rightarrow \hat{\boldsymbol{\beta}} =  (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}}}
\end{aligned}
$$

## Did we find a minimum?

$$
\begin{aligned}
\nabla^2_{\boldsymbol{\beta}} SSR &= \nabla_{\boldsymbol{\beta}} (-2\mathbf{X}^T\mathbf{y} + 2\mathbf{X}^T\mathbf{X}\boldsymbol{\beta}) \\[10pt]
&\class{fragment}{=-2\nabla_{\boldsymbol{\beta}}\mathbf{X}^T\mathbf{y} + 2\nabla_{\boldsymbol{\beta}}(\mathbf{X}^T\mathbf{X}\mathbf{\beta})} \\[10pt]
 &\class{fragment}{\propto \mathbf{X}^T\mathbf{X}}\class{fragment}{ > 0}
\end{aligned}
$$

<center><b>Show the details in HW 01</b></center>

# Predicted values and residuals

## Predicted (fitted) values

Now that we have $\hat{\boldsymbol{\beta}}$, let's predict values of $\mathbf{y}$ using the model

$$
\hat{\mathbf{y}} = \mathbf{X}\hat{\boldsymbol{\beta}} = \underbrace{\mathbf{X}(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T}_{\mathbf{H}}\mathbf{y} = \mathbf{H}\mathbf{y}
$$

. . .

**Hat matrix**: $\mathbf{H} = \mathbf{X}(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T$

. . .

-   $\mathbf{H}$ is an $n\times n$ matrix
-   Maps vector of observed values $\mathbf{y}$ to a vector of fitted values $\hat{\mathbf{y}}$
-   It is only a function of $\mathbf{X}$ not $\mathbf{y}$

## Residuals

Recall that the residuals are the difference between the observed and predicted values

$$
\begin{aligned}
\mathbf{e} &= \mathbf{y} - \hat{\mathbf{y}}\\[10pt]
&\class{fragment}{ = \mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}}} \\[10pt]
&\class{fragment}{ = \mathbf{y} - \mathbf{H}\mathbf{y}} \\[20pt]
\class{fragment}{\color{#993399}{\mathbf{e}}} &\class{fragment}{\color{#993399}{=(\mathbf{I} - \mathbf{H})\mathbf{y}}} \\[10pt]
\end{aligned}
$$

## Recap

-   Introduced matrix representation for simple linear regression

    -   Model from
    -   Least square estimate
    -   Predicted (fitted) values
    -   Residuals
