---
title: "ANOVA + Geometric interpretation"
author: "Prof. Maria Tackett"
date: "2024-09-17"
date-format: "MMM DD, YYYY"
footer: "[🔗 STA 221 - Fall 2024](https://sta221-fa24.netlify.app)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
filters:
  - parse-latex
bibliography: references.bib
---

## Announcements

-   Lab 02 due on **Thursday at 11:59pm**

    -   Push work to GitHub repo

    -   Submit final PDF on Gradescope + select all team members + mark pages for each question

-   HW 01 due **Thursday at 11:59pm**

    -   Note submission instructions

## Homework submission {.midi}

If you write your responses to Exercises 1 - 4 by hand, you will need to combine your written work to the completed PDF for Exercises 5 - 10 before submitting on Gradescope.

Instructions to combine PDFs:

-   Preview (Mac): [support.apple.com/guide/preview/combine-pdfs-prvw43696/mac](https://support.apple.com/guide/preview/combine-pdfs-prvw43696/mac)

-   Adobe (Mac or PC): [helpx.adobe.com/acrobat/using/merging-files-single-pdf.html](https://helpx.adobe.com/acrobat/using/merging-files-single-pdf.html)

    -   Get free access to Adobe Acrobat as a Duke student: [oit.duke.edu/help/articles/kb0030141/](https://oit.duke.edu/help/articles/kb0030141/)

## Latex in this class

For this class you will need to be able to...

-   Properly write mathematical symbols, e.g., $\beta_1$ not *B1,* $R^2$ not *R2*

-   Write basic regression equations, e.g., $\hat{y} = \beta_0 + \beta_1x_1 + \beta_2x_2$

-   Write matrix equations: $\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}$

-   Write hypotheses (we'll start this next week), e.g., $H_0: \beta = 0$

You are welcome to but <u>not</u> required to write math proofs using LaTex.

## Topics

-   Compare models using Adjusted $R^2$

-   Introduce the ANOVA table

-   Use a geometric interpretation to find the least squares estimates

## Computing setup

```{r packages}
#| echo: true
#| message: false

# load packages
library(tidyverse)
library(tidymodels)
library(openintro)
library(patchwork)
library(knitr)
library(kableExtra)
library(viridis) #adjust color palette

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))
```

## Data: Peer-to-peer lender

Today's data is a sample of 50 loans made through a peer-to-peer lending club. The data is in the `loan50` data frame in the **openintro** R package.

```{r}
#| echo: false

loan50 <- loan50 |>
  mutate(annual_income_th = annual_income / 1000)

loan50 |>
  select(annual_income_th, debt_to_income, verified_income, interest_rate)
```

## Variables

**Predictors**:

::: nonincremental
-   `annual_income_th`: Annual income (in \$1000s)
-   `debt_to_income`: Debt-to-income ratio, i.e. the percentage of a borrower's total debt divided by their total income
-   `verified_income`: Whether borrower's income source and amount have been verified (`Not Verified`, `Source Verified`, `Verified`)
:::

**Response**: `interest_rate`: Interest rate for the loan

## Model fit in R

```{r}
#| echo: true

int_fit <- lm(interest_rate ~ debt_to_income + verified_income  + annual_income_th, data = loan50)

int_fit2 <- lm(interest_rate ~ debt_to_income + verified_income  + annual_income_th + verified_income * annual_income_th, data = loan50)
```

# Model assessment and comparison

## RMSE & $R^2$

-   **Root mean square error, RMSE**: A measure of the average error (average difference between observed and predicted values of the outcome)

-   **R-squared**, $R^2$ : Percentage of variability in the outcome explained by the regression model

## Comparing models

::: incremental
-   Though we use $R^2$ to assess the model fit, it is generally unreliable for comparing models with different number of predictors. Why?

    -   $R^2$ will stay the same or increase as we add more variables to the model . Let's show why this is true.

    -   If we only use $R^2$ to choose a best fit model, we will be prone to choose the model with the most predictor variables.
:::

## Adjusted $R^2$

-   **Adjusted** $R^2$: measure that includes a penalty for unnecessary predictor variables
-   Similar to $R^2$, it is a measure of the amount of variation in the response that is explained by the regression model

## $R^2$ and Adjusted $R^2$

$$R^2 = \frac{SSM}{SST} = 1 - \frac{SSR}{SST}$$

<br>

. . .

$$R^2_{adj} = 1 - \frac{SSR/(n-p-1)}{SST/(n-1)}$$

where

-   $n$ is the number of observations used to fit the model

-   $p$ is the number of terms (not including the intercept) in the model

## Compare models {.midi}

Which model would you select `int_fit` (main effects only) or `int_fit2` (main effects + interaction) based on...

$R^2$

```{r}
glance(int_fit)$r.squared
glance(int_fit2)$r.squared
```

. . .

$Adj. R^2$

```{r}
glance(int_fit)$adj.r.squared
glance(int_fit2)$adj.r.squared
```

## ANOVA table {.midi}

```{=latex}
\begin{table}
\begin{tabular}{l|l|l|l|l}
 \hline
 Source & Sum of squares & DF & Mean square & F \\
 \hline
 Model & $\sum_{i=1}^n(\hat{y}_i - \bar{y})^2$ & $p$ & $SSM / p$ & $MSM / MSR$ \\ \hline
 Residual & $\sum_{i=1}^n(y_i- \hat{y}_i)^2$ & $n - p - 1$ & $SSR / (n - p - 1)$ & \\ \hline
 Total & $\sum_{i = 1}^n(y_i - \bar{y})^2$ & $n - 1$ & &  \\
 \hline
\end{tabular}
\end{table}
```
<br>

. . .

-   The **degrees of freedom (df)** are the number of independent pieces of information used to calculate a statistic.

-   **Mean square (MS)** is the sum of squares divided by the associated degrees of freedom.

## Using $R^2$ and Adjusted $R^2$

-   Adjusted $R^2$ can be used as a quick assessment to compare the fit of multiple models; however, it should not be the only assessment!

-   Use $R^2$ when describing the relationship between the response and predictor variables

# Geometric interpretation

## Geometry of least squares regression

::: incremental
-   Let $\text{Col}(\mathbf{X})$ be the **column space** of $\mathbf{X}$: the set all possible linear combinations (span) of the columns of $\mathbf{X}$

-   The vector of responses $\mathbf{y}$ is not in $\text{Col}(\mathbf{X})$.

-   **Goal:** Find another vector $\mathbf{z} = \mathbf{Xb}$ that is in $\text{Col}(\mathbf{X})$ and is as close as possible to $\mathbf{y}$.

    -   $\mathbf{z}$ is called a **projection** of $\mathbf{y}$ onto $\text{Col}(\mathbf{X})$ .
:::

## Geometry of least squares regression

::: incremental
-   For any $\mathbf{z} = \mathbf{Xb}$ in $\text{Col}(\mathbf{X})$, the vector $\mathbf{e} = \mathbf{y} - \mathbf{Xb}$ is the difference between $\mathbf{y}$ and $\mathbf{Xb}$.

    -   In other words, we want to minimize $||\mathbf{e}||^2 = ||\mathbf{y} - \mathbf{Xb}||^2$

-   This is minimized for the $\mathbf{b}$ ( we'll call it $\hat{\boldsymbol{\beta}}$ ) that makes $\mathbf{e}$ orthogonal to $\text{Col}(\mathbf{X})$

-   **Recall**: If $\mathbf{e}$ is orthogonal to $\text{Col}(\mathbf{X})$, then the inner product of any vector in $\text{Col}(\mathbf{X})$ and $\mathbf{e}$ is 0 $\Rightarrow \mathbf{X}^T\mathbf{e} = \mathbf{0}$
:::

## Geometry of least squares regression

-   Therefore, we have

$$
\mathbf{X}^T(\mathbf{y} - \mathbf{Xb}) = \mathbf{0}
$$

Let's solve for $\mathbf{b}$ to get the least squares estimate.

## Recap

-   Compared models using Adjusted $R^2$

-   Introduced the ANOVA table

-   Used a geometric interpretation to find the least squares estimates

## Next class

-   Inference for regression

-   See Sep 19 prepare
