---
title: " Variable transformations cont'd"
author: "Prof. Maria Tackett"
date: "2025-03-06"
date-format: "MMM DD, YYYY"
footer: "[🔗 STA 221 - Spring 2025](https://sta221-sp25.netlify.app)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 15-transformations-contd-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

## Announcements

-   HW 03 due March 20 at 11:59pm

-   Next project milestone: Exploratory data analysis due March 20

    -   Work on it in lab March 7

<br>

<center>Have a good spring break! 😎</center>

## Computing set up

```{r}
#| echo: true
#| message: false

# load packages
library(tidyverse)  
library(tidymodels)  
library(knitr)       
library(patchwork)

# set default theme in ggplot2
ggplot2::theme_set(ggplot2::theme_bw())
```

## Topics

-   Log-transformation on the predictor
-   Identify linear models

# Variable transformations

## Data: Life expectancy in 140 countries

```{r}
#| echo: false

library(readxl)

health_data <- read_xlsx("data/life-expectancy-data.xlsx") |>
  rename(life_exp = `Life_expectancy_at_birth`, 
         income_inequality = `Income_inequality_Gini_coefficient`,
         health_expenditure = Health_expenditure) |>
  mutate(education = if_else(Education_Index > median(Education_Index), "High", "Low"), 
         education = factor(education, levels = c("Low", "High")))

```

The data set comes from @zarulli2021 who analyze the effects of a country's healthcare expenditures and other factors on the country's life expectancy. The data are originally from the [Human Development Database](http://hdr.undp.org/en/data) and [World Health Organization](https://apps.who.int/nha/database/).

There are `r nrow(health_data)` countries (observations) in the data set.

::: aside
[Click here](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0253450) for the original research paper.
:::

## Variables

-   `life_exp`: The average number of years that a newborn could expect to live, if he or she were to pass through life exposed to the sex- and age-specific death rates prevailing at the time of his or her birth, for a specific year, in a given country, territory, or geographic income_inequality. ( from the [World Health Organization](https://www.who.int/data/gho/indicator-metadata-registry/imr-details/65#:~:text=Definition%3A,%2C%20territory%2C%20or%20geographic%20area.))

-   `income_inequality`: Measure of the deviation of the distribution of income among individuals or households within a country from a perfectly equal distribution. A value of 0 represents absolute equality, a value of 100 absolute inequality (based on Gini coefficient). (from @zarulli2021)

## Variables

-   `education`: Indicator of whether a country’s education index is above (`High`) or below (`Low`) the median index for the 140 countries in the data set.

    -   Education index: *Average of mean years of schooling (of adults) and expected years of school (of children), both expressed as an index obtained by scaling wit the corresponding maxima.*

-   `health_expend`: Per capita current spending on on healthcare goods and services, expressed in respective currency - international Purchasing Power Parity (PPP) dollar ([from the World Health Organization](https://www.who.int/data/gho/indicator-metadata-registry/imr-details/4952))

# Log transformation on a predictor variable

## Variability in life expectancy

Let's consider a model using a country's healthcare expenditure, income inequality, and education to predict its life expectancy

```{r}
#| echo: false
#| fig-align: center

p1 <- ggplot(data = health_data, aes(x = health_expenditure, y = life_exp)) + 
  geom_point() + 
  labs(x = "Healthcare expenditure",
       y = "Life expectancy")

p2 <- ggplot(data = health_data, aes(x = income_inequality, y = life_exp)) + 
  geom_point() + 
  labs(x = "Income inequality",
       y = "Life expectancy")

p3 <- ggplot(data = health_data, aes(x = education, y = life_exp)) + 
  geom_boxplot() + 
  labs(x = "Education",
       y = "Life expectancy")


p1 + p2 + p3
```

## Original model

```{r}
life_exp_fit <- lm(life_exp ~ health_expenditure + income_inequality + education, 
                   data = health_data)
```

```{r}
#| echo: false
tidy(life_exp_fit) |>
  kable(digits = 3)
```

## Original model: Residuals

```{r}
#| echo: false

life_exp_aug <- augment(life_exp_fit)

life_exp_resid <- ggplot(data = life_exp_aug, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.7) + 
  geom_hline(yintercept=0, color="red") +
  labs(x="Predicted", y="Residuals",
       title="Residuals vs. Predicted")

life_exp_resid
```

## Residuals vs. predictors

```{r}
#| echo: false
#| fig-align: center
resid1 <-  ggplot(data = life_exp_aug, aes(x = health_expenditure, y = .resid)) +
  geom_point(alpha = 0.7) + 
  geom_hline(yintercept=0, color="red") +
  labs(x="Health expenditure", y="Residuals",
       title="")

resid2 <-  ggplot(data = life_exp_aug, aes(x = income_inequality, y = .resid)) +
  geom_point(alpha = 0.7) + 
  geom_hline(yintercept=0, color="red") +
  labs(x="Income inequality", y="Residuals",
       title="")

resid3 <-  ggplot(data = life_exp_aug, aes(x = education, y = .resid)) +
  geom_boxplot() + 
  geom_hline(yintercept=0, color="red") +
  labs(x="Education", y="Residuals",
       title="")

resid1 + resid2 + resid3 + plot_annotation(title = "Residuals vs. predictors")


```

. . .

::: question
There is a non-linear relationship is between health expenditure and life expectancy.
:::

## Log Transformation on $X$

Try a transformation on $X$ if the scatterplot in EDA shows non-linear relationship and residuals vs. fitted looks parabolic

```{r}
#| echo: false
#| fig-align: center


p1 <- ggplot(data = health_data, aes(x = health_expenditure, y = life_exp)) + 
  geom_point() + 
  labs(x = "Healthcare expenditure",
       y = "Life expectancy",
       title = "EDA")

resid1 <-  ggplot(data = life_exp_aug, aes(x = health_expenditure, y = .resid)) +
  geom_point(alpha = 0.7) + 
  geom_hline(yintercept=0, color="red") +
  labs(x="Health expenditure", y="Residuals",
       title="Residuals")

p1 + resid1 + plot_annotation(title = "Health expenditure ")
```

## EDA

```{r}
#| echo: false
ggplot(data = health_data, aes(x = log(health_expenditure), y = life_exp)) + 
  geom_point() + 
  labs(x = "Log-transformed healthcare expenditure",
       y = "Life expectancy",
       title = "Life expectancy vs. Log-transformed health expenditure")
```

## Model with Transformation on $X_j$ {.midi}

When we fit a model with predictor $\log(X_j)$, we fit a model of the form

$$
\mathbf{Y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}, \quad \boldsymbol{\epsilon} \sim N(\mathbf{0}, \sigma^2_{\epsilon}\mathbf{I})
$$

such that $\mathbf{X}$ has a column for $\log(X_j)$ .

. . .

The estimated regression model is

$$
\begin{aligned}
\hat{\mathbf{Y}} &= \mathbf{X}\hat{\boldsymbol{\beta}} \\[8pt]
\Rightarrow \quad &\hat{y}_i = \hat{\beta}_0 + \hat{\beta}_1x_{i1} + \ldots + \hat{\beta}_j\log(x_{ij}) + \dots + \hat{\beta}_px_{ip} 
\end{aligned}
$$

## Model interpretation {.midi}

$$
\hat{y}_i = \hat{\beta}_0 + \hat{\beta}_1x_{i1} + \ldots + \hat{\beta}_j\log(x_{ij}) + \dots + \hat{\beta}_px_{ip} 
$$

::: incremental
-   **Intercept:** When $x_{i1} = \dots = \log(x_{ij}) = \dots = x_{ip} = 0$ , $y_i$ is expected to be $\hat{\beta}_0$, on average, holding all else constant.

    -   $\log(x_{ij}) = 0$ when $x_{ij} = 1$

-   **Coefficient of** $X_j$**:** When $x_{ij}$ is multiplied by a factor of $C$, $y_i$ is expected to change by $\hat{\beta}_1\log(C)$ units, on average, holding all else constant.

    -   **Example**: When $x_{ij}$ is multiplied by a factor of 2, $y_i$ is expected to increase by $\hat{\beta}_1\log(2)$ units, on average, holding all else constant.
:::

## Model with log(X) {.midi}

```{r}
life_exp_logx_fit <- lm(life_exp ~ log(health_expenditure) + income_inequality 
                        + education, data = health_data)

```

```{r}
#| echo: false

tidy(life_exp_logx_fit) |>
  kable(digits = 3)
```

<br>

::: question
-   Interpret the intercept in the context of the data.

-   Interpret the effect of health expenditure in the context of the data.

-   Interpret the effect of education in the context of the data.
:::

## Model with log(X): Residuals

```{r}
#| echo: false
life_exp_logx_aug <- augment(life_exp_logx_fit)

resid_logx <- ggplot(data = life_exp_logx_aug, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.7) + 
  geom_hline(yintercept=0, color="red") +
  labs(x="Predicted", y="Residuals",
       title="Residuals vs. Predicted")

resid_logx
```

## Comparing residual plots

```{r}

orig_resid <- ggplot(data = life_exp_aug, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.7) + 
  geom_hline(yintercept=0, color="red") +
  labs(x="Predicted", y="Residuals",
       title="Original Model: Residuals vs. Predicted")

logx_resid <- resid_logx <- ggplot(data = life_exp_logx_aug, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.7) + 
  geom_hline(yintercept=0, color="red") +
  labs(x="Predicted", y="Residuals",
       title="Transformed Model: Residuals vs. Predicted")

orig_resid + logx_resid
```

---

<center>

::: question
Is a model with log-transfromed response and/or predictor still a "linear" model?
:::

</center>

## “Linear” model

What does it mean for a model to be a “linear” model?

-   Linear models are linear in the parameters, i.e. given an observation $y_i$

    $$
    y_i = \beta_0 + \beta_1f(x_{i1}) + \dots + \beta_pf(x_{ip}) + \epsilon_i
    $$

    yi=β0+β1f1(xi1)+⋯+βpfp(xip)+ϵi

-   The functions $f_1, \ldots, f_p$ can be non-linear as long as $\beta_0, \beta_1, \ldots, \beta_p$ are linear in $Y$

## Identify the linear models

:::: extrapad
::: incremental
1.  $y_i = \beta_0 + \beta_1x_{i1} + \beta_2x_{i1}^2 + \beta_3x_{i2}  + \epsilon_i$

2.  $y_i = \beta_1x_{i1} + \beta_2x_{i2} + \beta_3x_{i1}x_{i2} + \epsilon_i$

3.  $y_i = \beta_0  + \beta_1\sin(x_{i1} + \beta_2x_{i2}) + \beta_3x_{i3} + \epsilon_i$

4.  $y_i = \beta_0 + \beta_1e^{x_{i1}} + \beta_2e^{x_{i2}} + \epsilon_i$

5.  $y_i =e^{(\beta_0 + \beta_1x_{i1} + \beta_2x_{i2} + \beta_3x_{i3})} + \epsilon_i$
:::
::::

## Learn more

See [Log Transformations in Linear Regression](https://github.com/sta210-sp20/supplemental-notes/blob/master/log-transformations.pdf) for more details about interpreting regression models with log-transformed variables.

## Recap

-   Introduced log-transformation on the predictor
-   Identified linear models

## Remaining questions? 

Please submit any questions you have about multicollinearity and variable transformations.

<center>

```{=html}
<iframe width="640px" height="480px" src="https://forms.office.com/r/QvBy4yvV8N?embed=true" frameborder="0" marginwidth="0" marginheight="0" style="border: none; max-width:100%; max-height:100vh" allowfullscreen webkitallowfullscreen mozallowfullscreen msallowfullscreen> </iframe>
```

</center>

## References
