---
title: "Logistic Regression: Inference"
author: "Prof. Maria Tackett"
date: "2024-11-14"
date-format: "MMM DD, YYYY"
footer: "[🔗 STA 221 - Fall 2024](https://sta221-fa24.netlify.app)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
    
bibliography: references.bib
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height =5,
	fig.width = 8,
	message = FALSE,
	warning = FALSE
)
```

## Announcements {.midi}

-   Project: Draft report due + peer review in December 2 lab

-   Statistics experience due Tuesday, November 26

-   HW 04 released later today. Due Thursday, November 21

## Topics

-   Test of significance for overall logistic regression model

-   Test of significance for a subset of model coefficients

-   Test of significance for a single coefficient

## Computational setup

```{r}
#| echo: true
#| warning: false
#| message: false


library(tidyverse)
library(tidymodels)
library(pROC)      
library(knitr)
library(kableExtra)

# set default theme in ggplot2
ggplot2::theme_set(ggplot2::theme_bw())
```

# 

## Risk of coronary heart disease {.midi}

This data set is from an ongoing cardiovascular study on residents of the town of Framingham, Massachusetts. We want to examine the relationship between various health characteristics and the risk of having heart disease.

-   `high_risk`:

    -   1: High risk of having heart disease in next 10 years
    -   0: Not high risk of having heart disease in next 10 years

-   `age`: Age at exam time (in years)

-   `totChol`: Total cholesterol (in mg/dL)

-   `currentSmoker`: 0 = nonsmoker, 1 = smoker

-   `education`: 1 = Some High School, 2 = High School or GED, 3 = Some College or Vocational School, 4 = College

```{r}
#| echo: false
#| message: false
#| warning: false


heart_disease <- read_csv("data/framingham.csv") |>
  select(age, education, TenYearCHD, totChol, currentSmoker) |>
  drop_na() |>
  mutate(
    high_risk = as_factor(TenYearCHD),
    education = as_factor(education),
    currentSmoker = as_factor(currentSmoker)
  )
```

## Modeling risk of coronary heart disease

Using `age`, `totChol`, `currentSmoker`

```{r}
heart_disease_fit <- glm(high_risk ~ age + totChol + currentSmoker, 
              data = heart_disease, family = "binomial")

tidy(heart_disease_fit, conf.int = TRUE) |> 
  kable(digits = 3)
```

# Test for overall significance

## Likelihood ratio test

Similar to linear regression, we can test the overall significance for a logistic regression model, i.e., whether there is at least one non-zero coefficient in the model

$$
\begin{aligned}
&H_0: \beta_1 = \dots = \beta_p = 0 \\
&H_a: \beta_j \neq 0 \text{ for at least one } j
\end{aligned}
$$

. . .

The **likelihood ratio test** compares the fit of a model with no predictors to the current model.

## Likelihood ratio test statistic

Let $L_0$ and $L_a$ be the likelihood functions of the model under $H_0$ and $H_a$, respectively. The **likelihood ratio test statistic** is

$$
G = -2[\log L_0 - \log L_a] = -2\sum_{i=1}^n \Big[ y_i \log \Big(\frac{\hat{\pi}^0}{\hat{\pi}^a_i}\Big) + (1 - y_i)\log \Big(\frac{1-\hat{\pi}^0}{1-\hat{\pi}^a_i}\Big)\Big]
$$

where $\hat{\pi}^0$ is the predicted probability under $H_0$ and $\hat{\pi}_i^a = \frac{\exp \{x_i^T\boldsymbol{\beta}\}}{1 + \exp \{x_i^T\boldsymbol{\beta}\}}$ is the predicted probability under $H_a$ [^1]

[^1]: See @wilks1935likelihood for explanation of why -2 is included.

## Likelihood ratio test statistic

$$
G = -2\sum_{i=1}^n \Big[ y_i \log \Big(\frac{\hat{\pi}^0}{\hat{\pi}^a_i}\Big) + (1 - y_i)\log \Big(\frac{1-\hat{\pi}^0}{1-\hat{\pi}^a_i}\Big)\Big]
$$

. . .

::: incremental
-   When $n$ is large, $G \sim \chi^2_p$, ( $G$ follows a Chi-square distribution with $p$ degrees of freedom)

-   The p-value is calculated as $\text{p-value} = P(\chi^2 > G)$

-   Large values of $G$ (small p-values) indicate at least one $\beta_j$ is non-zero
:::

## $\chi^2$ distribution

```{r}
#| echo: false
#| fig-height: 6

x <- seq(from =0, to = 10, length = 100)

# Evaluate the densities
y_1 <- dchisq(x, 1)
y_2 <- dchisq(x,2)
y_3 <- dchisq(x,3)
y_4 <- dchisq(x,5)

# Plot the densities
plot(x, y_1, col = 1, type = "l", ylab="",lwd=3, ylim = c(0, 0.5), 
     main  = "Chi-square Distribution")
lines(x,y_2, col = 2,lwd=3)
lines(x, y_3, col = 3,lwd=3)
lines(x, y_4, col = 4,lwd=3)

# Add the legend
legend("topright",
       c("df = 1", "df = 2 ", "df = 3", "df = 5"), 
       col = c(1, 2, 3, 4), lty = 1)
```

## Heart disease model: likelihood ratio test

$$
\begin{aligned}
&H_0: \beta_{age} = \beta_{totChol} = \beta_{currentSmoker} = 0 \\
&H_a: \beta_j \neq 0 \text{ for at least one }j
\end{aligned}$$

. . .

**Fit the null model**

(we've already fit the alternative model)

```{r}
#| echo: true

null_model <- glm(high_risk ~ 1, data = heart_disease, family = "binomial")

tidy(null_model) |>
  kable()
```

## Heart disease model: likelihood ratio test

**Calculate the log-likelihood for the null and alternative models**

```{r}
(L_0 <- glance(null_model)$logLik)
(L_a <- glance(heart_disease_fit)$logLik)
```

. . .

**Calculate the likelihood ratio test statistic**

```{r}
(G <- -2 * (L_0 - L_a))
```

. . .

## Heart disease model: likelihood ratio test

**Calculate the p-value**

```{r}
(p_value <- pchisq(G, df = 3, lower.tail = FALSE))
```

. . .

**Conclusion**

The p-value is small, so we reject $H_0$. The data provide evidence of at least one non-zero model coefficient in the model.

# Test a subset of coefficients

## Testing a subset of coefficients

-   Suppose there are two models:

    -   Reduced Model: includes predictors $x_1, \ldots, x_q$

    -   Full Model: includes predictors $x_1, \ldots, x_q, x_{q+1}, \ldots, x_p$

-   We can use the likelihood ratios to see if any of the new predictors are useful

. . .

$$
\begin{aligned}
&H_0: \beta_{q+1} = \dots = \beta_p = 0\\
&H_a: \beta_j \neq 0 \text{ for at least one }j
\end{aligned}
$$

. . .

This is called a **drop-in-deviance** test (also known as nested likelihood ratio test)

## Deviance

The **deviance** is a measure of the degree to which the predicted values are different from the observed values (compares the current model to a "saturated" model)

<br>

In logistic regression,

$$
D = -2 \log L 
$$

<br>

$D \sim \chi^2_{n - p - 1}$ ( $D$ follows a Chi-square distribution with $n - p - 1$ degrees of freedom)

<br>

Note: $n - p - 1$ a the degrees of freedom associated with the error in the model (like residuals)

## Drop-in-deviance test

$$
\begin{aligned}
&H_0: \beta_{q+1} = \dots = \beta_p = 0\\
&H_a: \beta_j \neq 0 \text{ for at least one }j
\end{aligned}
$$

. . .

The test statistic is

$$
\begin{aligned}
G = D_{reduced} - D_{full} &= -2\log L_{reduced} - (-2 \log L_{full}) \\
&= -2(\log L_{reduced} - \log L_{full})
\end{aligned}
$$

. . .

The p-value is calculated using a $\chi_{\Delta df}^2$ distribution, where $\Delta df$ is the number of parameters being tested (the difference in number of parameters between the full and reduced model).

## Heart disease model: drop-in-deviance test

Should we add `education` to the model?

-   Reduced model: `age`, `totChol`, `currentSmoker`

-   Full model: `age`, `totChol`, `currentSmoker` , `education`

. . .

$$
\begin{aligned}
&H_0: \beta_{ed1} = \beta_{ed2} = \beta_{ed3} = 0 \\
&H_a: \beta_j \neq 0 \text{ for at least one }j
\end{aligned}
$$

## Heart disease model: drop-in-deviance test {.midi}

```{r}
reduced_model <- glm(high_risk ~ age + totChol + currentSmoker, 
              data = heart_disease, family = "binomial")

full_model <- glm(high_risk ~ age + totChol + currentSmoker + education, 
              data = heart_disease, family = "binomial")
```

. . .

**Calculate deviances**

```{r}
(deviance_reduced <- -2 * glance(reduced_model)$logLik)
(deviance_full <- -2 * glance(full_model)$logLik)
```

. . .

**Calculate test statistic**

```{r}
(G <- deviance_reduced - deviance_full)
```

## Heart disease model: drop-in-deviance test

**Calculate p-value**

```{r}
pchisq(G, df = 3, lower.tail = FALSE)
```

<br>

. . .

::: question
What is your conclusion? Would you include `education` in the model that already has `age`, `totChol`, `currentSmoker`?
:::

## Drop-in-deviance test in R {.midi}

Conduct the drop-in-deviance test using the `anova()` function in R with option `test = "Chisq"`

```{r}
anova(reduced_model, full_model, test = "Chisq") |> 
  tidy() |> 
  kable(digits = 3)
```

## AIC and BIC

Similar to linear regression, we can use AIC and BIC to compare models.

$$
\begin{aligned}
&AIC = -2 \log L + 2(p+1) \\
&BIC = -2 \log L + \log(n)(p + 1)
\end{aligned}
$$

You want to select the model that minimizes AIC / BIC

## Compare models using AIC

AIC for reduced model (`age`, `totChol`, `currentSmoker`)

```{r}
glance(reduced_model)$AIC
```

<br>

. . .

AIC for full model (`age`, `totChol`, `currentSmoker`, `education`)

```{r}
glance(full_model)$AIC
```

## Compare models using BIC

BIC for reduced model (`age`, `totChol`, `currentSmoker`)

```{r}
glance(reduced_model)$BIC
```

<br>

. . .

BIC for full model (`age`, `totChol`, `currentSmoker`, `education`)

```{r}
glance(full_model)$BIC
```

# Test for a single coefficient

## Distribution of $\hat{\boldsymbol{\beta}}$

When $n$ is large, $\hat{\boldsymbol{\beta}}$, the estimated coefficients of the logistic regression model, is approximately normal.

<br>

::: question
How do we know the distribution of $\hat{\boldsymbol{\beta}}$ is normal for large $n$?
:::

## Distribution of $\hat{\boldsymbol{\beta}}$

The expected value of $\hat{\boldsymbol{\beta}}$ is the true parameter, $\boldsymbol{\beta}$, i.e., $E(\hat{\boldsymbol{\beta}}) = \boldsymbol{\beta}$

. . .

$Var(\hat{\boldsymbol{\beta}})$, the matrix of variances and covariances between estimators, is found by taking the second partial derivatives of the log-likelihood function (Hessian matrix)

$$
Var(\hat{\boldsymbol{\beta}}) = (\mathbf{X}^T\mathbf{V}\mathbf{X})^{-1}
$$

where $\mathbf{V}$ is a $n\times n$ diagonal matrix such that $V_{ii}$ is the estimated variance for the $i^{th}$ observation

## Test for a single coefficient

**Hypotheses:** $H_0: \beta_j = 0 \hspace{2mm} \text{ vs } \hspace{2mm} H_a: \beta_j \neq 0$, given the other variables in the model

. . .

**(Wald) Test Statistic:** $$z = \frac{\hat{\beta}_j - 0}{SE(\hat{\beta}_j)}$$

where $SE(\hat{\beta}_j)$ is the square root of the $j^{th}$ diagonal element of $Var(\hat{\boldsymbol{\beta}})$

. . .

**P-value:** $P(|Z| > |z|)$, where $Z \sim N(0, 1)$, the Standard Normal distribution

## Confidence interval for $\beta_j$

We can calculate the **C% confidence interval** for $\beta_j$ as the following:

$$
\Large{\hat{\beta}_j \pm z^* SE(\hat{\beta}_j)}
$$

where $z^*$ is calculated from the $N(0,1)$ distribution

. . .

::: callout-note
This is an interval for the change in the log-odds for every one unit increase in $x_j$
:::

## Interpretation in terms of the odds

The change in **odds** for every one unit increase in $x_j$.

$$
\Large{\exp\{\hat{\beta}_j \pm z^* SE(\hat{\beta}_j)\}}
$$

. . .

**Interpretation:** We are $C\%$ confident that for every one unit increase in $x_j$, the odds multiply by a factor of $\exp\{\hat{\beta}_j - z^* SE(\hat{\beta}_j)\}$ to $\exp\{\hat{\beta}_j + z^* SE(\hat{\beta}_j)\}$, holding all else constant.

## Coefficient for `age` {.midi}

```{r}
#| label: risk-model-age-highlight
#| echo: false

tidy(heart_disease_fit, conf.int = TRUE) |> 
  kable(digits = 3) |>
  row_spec(2, background = "#D9E3E4")
```

. . .

**Hypotheses:**

$$
H_0: \beta_{age} = 0 \hspace{2mm} \text{ vs } \hspace{2mm} H_a: \beta_{age} \neq 0
$$, given total cholesterol and smoking status are in the model.

## Coefficient for `age` {.midi}

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

**Test statistic:**

$$z = \frac{ 0.0825  - 0}{0.00575} = 14.34
$$

## Coefficient for `age` {.midi}

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

**P-value:**

$$P(|Z| > |14.34|) \approx 0
$$

. . .

```{r}
2 * pnorm(14.34,lower.tail = FALSE)
```

## Coefficient for `age` {.midi}

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

**Conclusion:**

The p-value is very small, so we reject $H_0$. The data provide sufficient evidence that age is a statistically significant predictor of whether someone is high risk of having heart disease, after accounting for total cholesterol and smoking status.

## CI for `age`

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

::: question
Interpret the 95% confidence interval for `age` in terms of the **odds** of being high risk for heart disease.
:::

## Overview of testing coefficients

**Test a single coefficient**

-   Likelihood ratio test

-   Drop-in-deviance test

-   Wald hypothesis test and confidence interval

. . .

**Test a subset of coefficients**

-   Likelihood ratio test

-   Drop-in-deviance test

. . .

Can use AIC and BIC to compare models in both scenarios

## References
