---
title: "AE 07: Newton Raphson"
date: "November 19, 2024"
editor: visual
bibliography: references.bib
---

::: callout-important
Go to the course GitHub organization and locate your ae-07 repo to get started.

Render, commit, and push your responses to GitHub by the end of class to submit your AE.
:::

```{r}
#| label: load-packages
#| warning: false
#| message: false

library(tidyverse)
library(knitr)
library(tidymodels)
```

# Introduction

## COVID-19 infection prevention practices at food establishments

Researchers at Wollo University in Ethiopia conducted a study in July and August 2020 to understand factors associated with good COVID-19 infection prevention practices at food establishments. Their study is published in @andualem2022.

They were particularly interested in the understanding implementation of prevention practices at food establishments, given the workers’ increased risk due to daily contact with customers.

We will use the data from @andualem2022 to explore the association between age, sex, years of service, and whether someone works at a food establishment with access to personal protective equipment (PPE) as of August 2020. We will use access to PPE as a proxy for wearing PPE.

The study participants were selected using a simple random sampling at the selected establishments.

.

```{r}
#| label: load-data
#| message: false

covid_df <- read_csv("data/covid-prevention-study.csv") |>
  rename(age = "Age of food handlers", 
         years = "Years of service", 
         ppe_access = "Availability of PPEs") |>
  mutate(sex = factor(if_else(Sex == 2, "Female", "Male"))) |>
  select(age, sex, years, ppe_access) |>
  mutate(ppe_access = as_factor(ppe_access))

```

## Functions for Newton-Raphson 

```{r}
## Calculate the first derivative of logL (score function)

calc_first_deriv <- function(beta, X, y){
  first_deriv <- rep(0, length(beta))
  for(i in 1:length(y)){
    first_deriv <- first_deriv + as.numeric(y[i] - exp(X[i,] %*% beta)/(1 + exp(X[i,] %*% beta))) %*% X[i,]
  }
  return(colSums(first_deriv)) #return values as a vector 
}

## Calculate the second derivative of logL (Hessian)

calc_second_deriv <- function(beta, X, y){
  second_deriv <- matrix(0, nrow = length(beta), ncol = length(beta))
  for(i in 1:length(y)){
    second_deriv <- second_deriv + as.numeric(1/(1 + exp(X[i,] %*% beta)))*
      as.numeric((exp(X[i,] %*% beta)/(1 + exp(X[i,] %*% beta)))) * 
      (X[i,]) %*% t(X[i,])
  }
  return(second_deriv)
}
```

## Get starting values 

```{r}

# design matrix
X <- model.matrix(~ age + sex + years, data = covid_df)

# vector of response
y <- I(covid_df$ppe_access == 1)

# vector of coefficients
beta <- c(0, 0, 0, 0)

# keep track of iterations
iter <- 1

# keep track of difference in estimates
delta <- 1

# keep track of estimates at each iteration 
temp <- matrix(0, nrow = 500, ncol = 4) 
```

##  Estimate $\boldsymbol{\beta}$ 

```{r}
while(delta > 0.000001 & iter < 50){
  old <- beta
  beta <- old - solve(-1 * calc_second_deriv(beta = beta, X = X, y = y)) %*% 
                calc_first_deriv(beta = beta, X = X, y = y)
  temp[iter,] <- beta
  delta <- sqrt(sum((beta - old)^2))
  iter <- iter + 1
}
```

## Show results 

```{r}
iter
beta
```

Note that $(\nabla^2_{\boldsymbol{\beta}} \log L)^{-1}$is the estimate for $Var(\boldsymbol{\hat{\beta}})$ . The square root of the diagonal elements are the estimates for $SE(\hat{\boldsymbol{\beta}})$.

```{r}

#calculate the hessian matrix
second_deriv <- calc_second_deriv(beta = beta, X = X, y = y)
  
# take the inverse
inv_second_deriv <- solve(second_deriv)

# get estimates for SE
se_beta <- sqrt(diag(inv_second_deriv))

se_beta
```

## Coefficient estimates from `glm`

```{r}
ppe_model <- glm(ppe_access ~ age + sex + years, 
                 data = covid_df, family = binomial)
tidy(ppe_model, conf.int = TRUE) |>
  kable(digits = 3)
```

## References
