---
title: "AE 01: Simple linear regression "
subtitle: "Houses in Duke Forest"
date: "Sep 3, 2024"
editor: visual
toc: true
execute: 
  message: false
  warning: false
---

::: callout-important
Go to the [course GitHub organization](https://github.com/sta221-fa24) and locate your `ae-01` repo to get started.

Render, commit, and push your responses to GitHub by the end of class to submit your AE.\
\
**This AE will not count towards your participation grade.**
:::

```{r}
#| label: load-packages
#| message: false

library(tidyverse)    # data wrangling and visualization
library(tidymodels)   # broom and yardstick package
library(openintro)    # duke_forest dataset
library(knitr)        # format output
library(scales)       # format plot axes
library(skimr)        # quickly calculate summary statistics
```

# Data

The data are on houses that were sold in the Duke Forest neighborhood of Durham, NC around November 2020. It was originally scraped from Zillow, and can be found in the `duke_forest` data set in the **openintro** R package.

We will focus on two variables:

-   `area`: Total area of the home in square feet (sqft)

-   `price`: Sale price in US Dollars (USD)

The goal of this analysis is to use the area to understand variability in the price of homes in Duke Forest.

```{r}
#| label: glimpse-data

glimpse(duke_forest)
```

# Exploratory data analysis

Let's begin by examining the univariate distributions of the price and area. The code to visualize and calculate summary statistics for `price` is below.

```{r}
#| label: price-viz

ggplot(data = duke_forest, aes(x = price)) + 
  geom_histogram() +
  labs(x = "Price in US dollars", 
       title = "Price of houses in Duke Forest") + 
  scale_x_continuous(labels = label_dollar(scale_cut = cut_long_scale()))
```

```{r}
#| label: price-summary
duke_forest |>
  summarise(min = min(price), q1 = quantile(price, 0.25), 
            median = median(price), q3 = quantile(price, 0.75), 
            max = max(price), mean = mean(price), sd = sd(price)) |>
  kable(digits = 3)
```

## Exercise 1

What are 1 - 2 observations about the distribution of `price`?

## Exercise 2

Visualize the distribution of `area` and calculate summary statistics.

```{r}
#| label: ex2-area-viz

# add code here

```

```{r}
#| label: ex2-area-summary

# add code here

```

## Exercise 3

What are 1 - 2 observations about the distribution of `area`?

## Exercise 4

Fill in the code to visualize the relationship between `price` and `area`. What are 1 - 2 observations about the relationship between these two variables?

::: callout-important
Remove `#|eval: false` after you have filled in the code!
:::

```{r}
#| label: scatterplot
#| eval: false

ggplot(duke_forest, aes(x = ____, y = ____)) +
  geom_point(alpha = 0.7) +
  labs(
    x = "_______",
    y = "_________",
    title = "Price and area of houses in Duke Forest"
  ) +
  scale_y_continuous(labels = label_dollar()) 
```

# Regression model

## Exercise 5

You want to fit a model of the form

$$
price = \beta_0 + \beta_1 ~ area + \epsilon, \hspace{5mm} \epsilon \sim N(0, \sigma^2_\epsilon)
$$

Would a model of this form be a reasonable fit for the data? Why or why not?

## Exercise 6

Fit the linear model described in the previous exercise and neatly display the output.

See [notes](https://sta221-fa24.netlify.app/slides/02-slr#/linear-regression-in-r) for example code.

```{r}
#| label: fit-model

# add code here

```

## Exercise 7

-   Interpret the slope in the context of the data.

-   Interpret the slope in terms of area increasing by 100 sqft.

-   Which interpretation do you think is more meaningful in practice?

## Exercise 8

Does it make sense to interpret the intercept? If so, interpret it in the context of the data. Otherwise, explain why not.

# Submission

::: callout-important
To submit the AE:

-   Render the document to produce the PDF with all of your work from today's class.
-   Push all your work to your AE repo on GitHub. You're done! 🎉
:::
