---
title: "AE 02: Multiple linear regression"
subtitle: "Peer-to-peer lending"
date: "Sep 12, 2024"
editor: visual
---

::: callout-important
Go to the [course GitHub organization](https://github.com/sta221-fa24) and locate your `ae-02` repo to get started.

Render, commit, and push your responses to GitHub by the end of class to submit your AE.
:::

# Packages

```{r}
#| label: load-packages
#| message: false

library(tidyverse)   
library(tidymodels)   
library(openintro)    
library(knitr)       
```

# Data

Today's data is a sample of 50 loans made through a peer-to-peer lending club. The data is in the `loan50` data frame in the **openintro** R package.

We will focus on the following variables:

-   `annual_income_th`: Annual income (in \$1000s)
-   `debt_to_income`: Debt-to-income ratio, i.e. the percentage of a borrower's total debt divided by their total income
-   `verified_income`: Whether borrower's income source and amount have been verified (`Not Verified`, `Source Verified`, `Verified`)
-   `interest_rate`: Interest rate for the loan

The goal of this analysis is to use the annual income, debt-to-income ratio, and income verification to understand variability in the interest rate on the loan.

We'll start with data prep to rescale annual income to \$1000's and recode `verified_income` to fix an issue with the underlying data.

```{r}
#| label: data-prep

loan50 <- loan50 |>
   mutate(annual_income_th = annual_income / 1000, 
          verified_income = 
            case_when(verified_income == "Not Verified" ~ "Not Verified",
                      verified_income == "Source Verified" ~ "Source Verified",
                      verified_income == "Verified" ~ "Verified"),
          verified_income = as_factor(verified_income)
   )                    
```

```{r}
#| label: glimpse-data

glimpse(loan50)
```

# Categorical predictors

::: callout-note
## Exercise 1

Let's take a look at the design matrix for the model with predictors `debt_to_income`, `annual_income_th`, and `verified_income`.

How does R choose the baseline level by default?
:::

```{r}
#| label: ex-1

## add code here

```

\[Add response here\]

::: callout-note
## Exercise 2

Fit the model with the predictors `debt_to_income`, `annual_income_th`, `verified_income` , and the interaction between `annual_income_th` and `verified_income`.

Neatly display the model results using 3 digits.
:::

```{r}
#| label: ex-2

# add code here

```

::: callout-note
## Exercise 3

-   Write the estimated regression equation for the people with `Not Verified` income.

-   Write the estimated regression equation for people with `Verified` income.
:::

\[add response here\]

::: callout-note
## Exercise 4

In general, how do

-   indicators for categorical predictors impact the model equation?

-   interaction terms impact the model equation?
:::

\[Add response here\]

# Model assessment

::: callout-note
## Exercise 5

Let's compare the original model without interaction effects to the model you fit in Exercise 2.

Calculate $R^2$ and $Adj. R^2$ for each model. You can find $Adj. R^2$ from the `glance` function:

`glance(model_name)$adj.r.squared`
:::

```{r}
#| label: original-model 

int_fit <- lm(interest_rate ~ debt_to_income + annual_income_th +
                verified_income, data = loan50)
```

```{r}
#| label: ex-5

# add code here

```

::: callout-note
## Exercise 6

Which model would you choose based on

-   $R^2$?

-   $Adj. R^2$?
:::

\[add response here\]

# LaTex

Sometimes, you will need to include mathematical notation in your document. There are two ways you can display mathematics in your document:

**Inline:** Your mathematics will display within the line of text.

-   Use `$` to start and end your LaTex syntax. You can also use the menu: *Insert -\> LaTex Math -\> Inline Math*.

-   *Example*: The text `The simple linear regression model is $\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}$` produces\

    The simple linear regression model is $\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}$

\
**Displayed:** Your mathematics will display outside the line of text

-   Use a `$$` to start and end your LaTex syntax. You can also use the menu: *Insert -\> LaTex Math -\> Display Math*.

-   *Example*: The text `The estimated regression equation is $$\hat{\mathbf{y}} = \mathbf{X}\hat{\boldsymbol{\beta}}$$`produces\

    The estimated regression equation is

$$
\hat{\mathbf{y}} = \mathbf{X}\hat{\boldsymbol{\beta}}
$$

::: callout-tip
[Click here](https://oeis.org/wiki/List_of_LaTeX_mathematical_symbols) for a quick reference of LaTex code.
:::

# Submission

::: callout-important
To submit the AE:

-   Render the document to produce the PDF with all of your work from today's class.
-   Push all your work to your AE repo on GitHub. You're done! 🎉
:::
