---
title: "AE 03: Inference"
subtitle: "NCAA Football Expenditures"
date: "Sep 19, 2024"
editor: visual
---

::: callout-important
Go to the course GitHub organization and locate your ae-03 repo to get started.

Render, commit, and push your responses to GitHub by the end of class to submit your AE.
:::

# Set up

```{r}
#| label: load-packages-data
#| message: false

library(tidyverse)
library(tidymodels)
library(knitr)

football <- read_csv("data/ncaa-football-exp.csv")
```

# Data

# Regression model

```{r}
exp_fit <- lm(total_exp_m ~ enrollment_th + type, data = football)

tidy(exp_fit)|> 
  kable(digits = 3)
```

# Hypothesis test

We want to conduct a hypothesis test to determine if there is a linear relationship between enrollment and football expenditures after accounting for institution type.

We'll start by getting estimates for statistics we'll need for inference.

::: callout-note
## Exercise 1

We will use the vector of responses $\mathbf{y}$ and the design matrix $\mathbf{X}$ to calculate the values needed for inference.

Get $\mathbf{y}$ and $\mathbf{X}$ from the football data frame. What are their dimensions?
:::

```{r}
#| label: ex-1

# add code here

```

::: callout-note
## Exercise 2

Next, let's calculate $\hat{\sigma}_\epsilon^2$ the estimate. Use $\mathbf{y}$ and $\mathbf{X}$ from the previous exercise to calculate this value.
:::

```{r}
#| label: ex-2

## add code here

```

::: callout-note
## Exercise 3

Now we're ready to conduct the hypothesis test. State the null and alternative hypotheses in words and using mathematical notation.
:::

. . .

::: callout-note
## Exercise 4

Calculate $SE(\beta_j)$, then use this value to calculate the test statistic for the hypothesis test.
:::

```{r}
#| label: ex-4


## add code here

```

::: callout-note
## Exercise 5

Now we need to calculate p-value to help make our final conclusion.

-   State the distribution used to calculate the p-value.

-   Fill in the code below to calculate the p-value. Remove `#| eval: false` once you've filled in the code.
:::

```{r}
#| label: ex-5
#| eval: false

pt([test-statistic], [df], lower.tail = FALSE)

```

::: callout-note
## Exercise 6

State your conclusion in the context of the data. Use a threshold of $\alpha = 0.05$.
:::

. . .

::: callout-important
# Submission

To submit the AE:

Render the document to produce the PDF with all of your work from today's class.

Push all your work to your AE repo on GitHub. You're done! 🎉
:::
